---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mrQC

<!-- badges: start -->
<!-- badges: end -->

mrQC is an R package for checking the accuracy of meta- and summary-data prior to their use in Mendelian randomisation analyses. For example, the package provides tools for checking that the reported effect allele and effect allele frequency columns are correct. It also checks for possible errors in the summary data that might introduce bias into downstream Mendelian randomisation analyses.   

## Installation

To install the latest version of mrQC, perform as normal:

``` r
install.packages("devtools")
devtools::install_github("MRCIEU/mrQC")
```

## Example. Check that effect allele column is correct

First retrieve the top hits from the NHGRI-EBI GWAS catalog for the trait of interest. In this example our trait of interest is head and neck cancer. We recommend that you search on the trait's EFO, as this should generally identify more hits (compared to searching on the trait description/name).  

```{r example_hnc}
library(mrQC)
snplist<-make_snplist(efo_id="EFO_0004541")
length(snplist)
head(snplist)
```

We see that our search identified 187 genetic associations for head and neck cancer in the NHGRI-EBI GWAS catalog. 

