---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mrQC

<!-- badges: start -->
<!-- badges: end -->

mrQC is an R package for checking the accuracy of meta- and summary-data for outcome traits prior to their use in Mendelian randomisation analyses. For example, the package provides tools for checking that the reported effect allele and effect allele frequency columns are correct. It also checks for possible errors in the summary data that might introduce bias into downstream Mendelian randomisation analyses.   

## Installation

To install the latest version of mrQC, perform as normal:

``` r
install.packages("devtools")
devtools::install_github("MRCIEU/mrQC")

```
## General overview 

This package exploits three groups of SNPs: 1) **the MAF 1KG reference set**. This a set of 2297 SNPs that have the same minor allele across the 1000 genomes super populations and that have a minor allele frequency between 0.1 and 0.3. These SNPs are used to check for allele frequency errors; 2) **the GWAS catalog top hits**. These are SNPs in the GWAS catalog that are strongly associated with the outcome trait of interest; and 3) the **exposure SNPs** or the genetic instruments for the exposure of interest. 

The package performs 4 types of checks: 

1) confirm the identity of the effect allele column. This step relies on comparison of GWAS catalog top hits.  
2) confirm the identity of the effect allele frequency column. This step relies on comparison of the MAF reference set.  
3) assess the ancestral origins of a study. This step relies on comparison of the MAF reference set.  
4) for case-control studies, confirm that the provided effect sizes are log odds ratios or identify potential summary data errors. This step can use any of the SNP sets. 

## Step 1. Extract and format the target data

The first step is to make a list of SNP rsids. Let's assume our outcome trait of interest is glioma. Let's identify the top GWAS hits for glioma in the GWAS catalog. We will search on both the reported trait name as well as the trait experimental factor ontology (EFO). Note that not all studies in the GWAS catalog have up-to date EFO annotations. It is therefore advisable to search on both efo and reported trait, to maximise the number of retrieved hits. 


```{r make_snplist1, eval = TRUE}
library(mrQC)
snplist<-make_snplist(efo="glioma",trait="glioma",ref1000G_superpops=FALSE) #search on EFO abd reported trait 
head(snplist)
length(snplist)
```
We see that searching on the glioma EFO and glioma trait retrieves 54 unique SNP rsids. We could also have searched on the glioma efo_id, which retrieves the same SNPs as searching on the "glioma" efo.

```{r make_snplist2, eval = FALSE}
snplist<-make_snplist(efo_id="EFO_0005543",ref1000G_superpops=FALSE) 
```

In the above examples we set ref1000G_superpops to FALSE. We now set this to TRUE, which will result in a SNP list that includes the "MAF reference set". This is a set of 2297 SNPs that have the same minor allele across all 1000 genomes super populations and a minor allele frequency of 0.1-0.3.  


```{r make_snplist3, eval = TRUE}
snplist<-make_snplist(efo="glioma",trait="glioma",ref1000G_superpops=TRUE) 
head(snplist)
length(snplist)
```

We can also define a set of "exposure SNPs" and include this in list of rsids. For example, let's assume we are conducting a Mendelian randomisation study to assess the effect of polyunsaturated fatty acid exposure on risk of glioma. Let's define the "exposure SNPs", or genetic instrument, as SNPs associated with polyunsaturated fatty acids with a P value <5e-8. Let's use the ieugwasr package to extract "exposure SNPs" for polyunsaturated fatty acids from the Open GWAS project and also add these to the other SNPs. 

```{r make_snplist4, eval = TRUE}
instruments<-ieugwasr::tophits(id="met-d-PUFA",pval = 5e-08)
snplist<-make_snplist(efo = "glioma",trait="glioma",ref1000G_superpops=TRUE,snplist_user=instruments$rsid)
head(snplist)
length(snplist)
```

Our SNP list now contains the rsids for: 1) the GWAS catalog top hits, 2) the MAF reference set and 3) the "exposure SNPs". Next, we extract the summary associations statistics for these SNPs from the glioma outcome dataset. The example below has already been restricted to the SNPs of interest to save space. The extract_snps function only works on MAC or linux operating systems. 

```{r extract_snps1, eval = TRUE}
File<-system.file("extdata", "glioma_test_dat.txt", package = "mrQC")
gli<-extract_snps(snplist=snplist,path_to_target_file=File,path_to_target_file_sep="\t")
dim(gli)
head(gli)
```

In the above example, the summary data for glioma was stored locally on our machine in a tab separated text file. Alternatively, we could have sourced the outcome summary data from the Open GWAS project (https://gwas.mrcieu.ac.uk/). For example, to extract summary data for thyroid cancer we could run:

```{r extract_snps2,eval = TRUE}
snplist<-make_snplist(efo = "thyroid carcinoma",trait="thyroid carcinoma",ref1000G_superpops=TRUE,snplist_user=instruments$rsid)
head(snplist)
length(snplist)
thy <- ieugwasr::associations(id="ieu-a-1082", variants=snplist,proxies=0)  
dim(thy)
thy
```

The make_snplist() function returns a warning that no GWAS hits were found when searching for thyroid carcinoma as the reported trait. However the search was able to identify hits searching on the efo for thyroid carcinoma. 

Returning to the glioma example, having extracted the summary data for the SNP rsids of interest, we now need to format the summary data. This is to get the data into the expected format for the QC functions. 

```{r format_data1,eval = TRUE}
Dat<-format_data(dat=gli,outcome="Glioma",population="European",pmid=22886559,study="GliomaScan",ncase="cases",ncontrol="controls",UKbiobank=FALSE,rsid="Locus",effect_allele="Allele1",other_allele="Allele2",or="OR",lci="OR_95._CI_l",uci="OR_95._CI_u",eaf="eaf.controls",p="p",efo="glioma")
head(Dat)
```

Now we are ready to perform some quality checks on the summary data

## Step 2. Check that the effect allele frequency column is correct

Next we create some plots to visualise potential problems with the effect allele frequency column. We do this by comparing allele frequency in the outcome glioma dataset to the 1000 genomes super populations. Let's restrict the comparison to the European super population, since we know that the glioma dataset was derived from a European ancestry population. 

```{r make_maf_plot1,eval = TRUE}
Plot1<-make_plot_maf(snp_target="rsid",ref_1000G="EUR",target_dat=Dat,target_dat_population="population",target_study="study",target_dat_effect_allele="effect_allele",target_dat_other_allele="other_allele")
Plot1
```

SNPs with a red colour are SNPs with incompatible minor allele frequencies, i.e. the allele frequencies are above 0.5 in the target dataset but less than 0.5 in the 1000 genomes dataset. In this example, all SNPs are flagged as problematic and there is a strong inverse correlation in minor allele frequency between the datasets. This indicates that the reported effect allele frequency column, in the outcome dataset, does not correspond to the reported effect allele. The strong inverse correlation implies that the effect allele column actually refers to the non-effect allele. 

Next we construct a similar plot but this time comparing allele frequency with all 1000 genomes super populations.

```{r make_maf_plot2,eval = FALSE}
Plot2<-make_plot_maf(snp_target="rsid",ref_1000G=c("AFR","AMR", "EAS", "EUR", "SAS","ALL"),target_dat=Dat,target_dat_population="population",target_study="study",target_dat_effect_allele="effect_allele",target_dat_other_allele="other_allele")
Plot2
```


All SNPs across all super populations are flagged as problematic. However, we only see a strong inverse correlation in the comparison with the European 1000 genomes super population because the glioma example has been derived from a European ancestry population. This example serves to illustrate that our method can identify problematic SNPs regardless of the ancestry of the outcome dataset. By assessing the strength of the correlation in MAF between the datasets, once can also make inferences about the ancestral background of the outcome dataset (assuming one didn't know this already). 

In the next example, we compare effect alleles between the glioma outcome dataset and the GWAS catalog. 

## Step 3. Check that the effect allele column is correct


```{r make_gwascatalog_plot1,eval = FALSE}
Plot3<-make_plot_gwas_catalog(dat=Dat,efo=unique(Dat$efo),trait="glioma")
Plot3
```

We see that there are three groups of SNPs: those flagged as showing no effect size conflict; those with moderate effect size conflict; and those with high effect size conflict. The distinction between moderate and high effect size conflict is arbitrary but is specified to make allowance for chance deviations between datasets. If the effect sizes are in opposite directions, the effect size conflict flag is set to moderate or high. If the two-sided P value in both datasets is ≤0.0001 then the flag is upgraded to high. In addition, if the summary association statistics in the target dataset and GWAS catalog are derived from the same publication, and effect sizes are in opposite directions, the conflict flag is set to high regardless of P value. 

Overall, the plot shows that the majority of SNPs show effect size conflicts. In other words, SNPs associated with higher risk of glioma in the GWAS catalog tend to be associated with lower risk of glioma in the target dataset. This indicates that the reported effect allele column may actually correspond to the non-effect allele. Taken together with the previous allele frequency plots, showing an inverse correlation in MAF between the target dataset and 1000 genomes European superpopulation, this strongly indicates that the effect and non-effect allele columns have been incorrectly specified. The reported non-effect allele column is actually the effect allele column.

We can also make a plot comparing effect allele frequency between the target dataset and the GWAS catalog, which we show in the next example.

```{r make_gwascatalog_plot2,eval = FALSE}

Plot4<-make_plot_gwas_catalog(dat=Dat,plot_type="plot_eaf",efo=unique(Dat$efo),trait=unique(Dat$outcome))
```

We see an inverse correlation in effect allele frequency (EAF) between the target dataset and the GWAS catalog in European ancestry studies. This is the expected pattern when the effect allele column has been incorrectly specified and reflects the non-effect allele. The EAF conflict flag is set to moderate or high if EAF is not consistently greater or less than 0.5 in the target dataset and the GWAS catalog (e.g. is <0.5 in the target dataset but is >0.5 in the GWAS catalog). For conflicting SNPs, the flag is upgraded to high if effect allele frequency is >0.6 or <0.4. This makes allowance for chance deviations in allele frequency for SNPs with MAF close to 0.5.   

In the next example, we compare the predicted log odds ratio to the reported effect size. This step is applicable to summary data that has been derived from a logistic regression model of case-control status. Since the function to derive the predicted log odds ratio can be a bit slow, we restrict the glioma example to just the first 20 SNPs. 


## Step 4. Check whether the reported effect size corresponds to log odds ratios 

```{r make_plot_predlnor1,eval = FALSE}
Dat1<-Dat[1:20,]
Pred<-predict_lnor_sh(dat=Dat1)
Plot5<-make_plot_predlnor(dat=Pred)
Plot5
```
The plot shows a strong positive correlation between the predicted and reported log odds ratio. This is expected. What's more useful is the intercept and the slope, which are displayed as part of the figure subheading. In this example, the intercept is close to zero and the slope is >0.8, which is close to the expected values. When the predicted and reported log odds ratios are identical, the intercept should be 0 and the slope should be 1. 


We can also plot the bias, i.e. the deviation of the predicted log odds ratio from the reported effect size. 

```{r make_plot_predlnor3,eval = FALSE}
Plot6<-make_plot_predlnor(dat=Pred,bias=TRUE)
Plot6
```
Overall the bias seems small and mostly varies from -10.9% to -13.5%,  although it does seem to get worse with stronger effect sizes. Since genetic effect sizes tend to be quite small, a bias of 10% is unlikely to make a material difference. 

In the next example we show a dataset that returns a slope and intercept very different from expectation

```{r make_plot_predlnor2,eval = FALSE}
snplist<-make_snplist(efo = "kidney cancer",snplist_user=instruments$rsid)
ukb <- data.frame(ieugwasr::associations(id="ukb-b-1316", variants=snplist,proxies=0),stringsAsFactors=FALSE)
Ukb<-format_data(dat=ukb,outcome="Kidney cancer",population="European",pmid="ukb-b-1316",ncase=1114,ncontrol=461896,study="UKB",UKbiobank=TRUE,rsid="rsid",effect_allele="ea",other_allele="nea",lnor="beta",se="se",eaf="eaf",p="p",effect_allele_confirmed=TRUE,all_summary_stats=TRUE,ID=145,efo = "kidney cancer")
Pred<-predict_lnor_sh(dat=Ukb)
Plot5_2<-make_plot_predlnor(dat=Pred)
Plot5_2
```
Unsurprisingly there is a strong positive correlation. However, the slope is 400, when we expect a slope of 1. In fact, further investigation reveals that Open GWAS dataset ukb-b-1316 has been generated using a linear mixed model. In other words, the results were derived from a model where kidney cancer case-control status (controls coded 1 and cases coded 2) was regressed on SNP genotype (additively coded). The effect size from this model can be interpreted as the absolute risk of kidney cancer per copy of the  effect allele. We can transform this into a log odds ratio scale using the transform_betas() function.  

```{r make_plot_predlnor4,eval = FALSE}
Ukb2<-transform_betas(dat=Ukb,effect="lnor",effect.se="se")
Pred2<-predict_lnor_sh(dat=Ukb2)
Plot5_3<-make_plot_predlnor(dat=Pred2)
Plot5_3
```

After transforming the reported effect size to a log odds ratio scale, we now see a slope close to 1 for the relationship with the predicted log odds ratio. We suggest that any dataset with an unsual intercept or with a slope very different from 1 (e.g. <0.8 or >1.2) should be investigated by the user for potential problems. In the glioma example, the slope and intercept look reasonably close to what we'd expect. In the kidney cancer example, the slope was different from 1 because the reported effect sizes had not been generated in a logistic regression model. Other factors that could cause the slope to differ from 1 include: the impact of covariate adjustment in the original GWAS, deviations from Hardy Weinberg equilibrium, mismatches between reported and actual allele frequency or mismatches between reported and actual SNP-level sample size.    

Let's now return back to the glioma dataset. In the next example we generate ZZ plots, which can be used to flag SNPs with P values that don't coincide with their reported effect sizes. The zz_plot() function compares Zp scores (inferred from the reported P values) to Zb scores (inferred from the reported effect size and standard error).

## Step 5. Check whether the P values correspond to the effect sizes  

```{r make_zz_plot1,eval = FALSE}
Plot7<-zz_plot(dat=Dat)
Plot7
```

The correlation between the Zp and Zb scores is 1, indicating very strong concordance between the reported P values and reported effect sizes in the glioma dataset for the selected SNPs. In the next example, we highlight a dataset where the correlation is less than 1. 

```{r make_zz_plot2,eval = FALSE}
instruments<-ieugwasr::tophits(id="met-d-PUFA")
snplist<-make_snplist(efo = "lung carcinoma",snplist_user=instruments$rsid)
luc <- ieugwasr::associations(id="ieu-a-966", variants=snplist,proxies=0)  
Dat<-format_data(dat=data.frame(luc,stringsAsFactors=F),outcome="Lung cancer",population="European",pmid=24880342,ncase=11348,ncontrol=15861,study="ILCCO",UKbiobank=FALSE,rsid="rsid",effect_allele="ea",other_allele="nea",lnor="beta",se="se",eaf="eaf",p="p",efo = "lung carcinoma")
Plot7_2<-zz_plot(dat=Dat)
Plot7_2
```

The correlation between the Zp and Zb scores is less than 1, suggesting discordance between the reported P values and reported effect sizes. In the example, we included all three sets of SNP (the GWAS catalog hits, the MAF reference set and the "exposure SNPs". Let's restrict the comparison to only the "exposure SNPs" and the GWAS catalog top hits. 


```{r make_zz_plot3,eval = FALSE}
snplist<-make_snplist(efo = "lung carcinoma",snplist_user=instruments$rsid,ref1000G_superpops=FALSE)
luc <- ieugwasr::associations(id="ieu-a-966",variants=snplist,proxies=0)  
Dat<-format_data(dat=data.frame(luc,stringsAsFactors=F),outcome="Lung cancer",population="European",pmid=24880342,ncase=11348,ncontrol=15861,study="ILCCO",rsid="rsid",effect_allele="ea",other_allele="nea",lnor="beta",se="se",eaf="eaf",p="p",efo = "lung carcinoma")
Plot7_3<-zz_plot(dat=Dat)
Plot7_3
```

The correlation between the Zp and Zb scores is still less than 1, suggesting some discordance between the reported P values and reported effect sizes. For example, 1 SNP has a Z score close to 15 when estimated from the P value but a Z score close to 5 when estimated from the reported effect size and standard error 


## Step 6. Combine all plots into a single report 

Next we combine all the plots into a single report that we can save for later. 

```{r combine_plots,eval = FALSE}
Plot_list2<-ls()[grep("Plot[0-9]",ls())] 
Plot_list<-lapply(1:length(Plot_list2),FUN=function(x) eval(parse(text=Plot_list2[x])))
combine_plots(Plot_list=Plot_list,out_file="~/qc_report.png")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/qc_report.png"
```



